### 2019-08-28 Manuscript/Manuscript.tex

Of the various parameters used in the equations that govern the silicate weathering component of GEOCLIM (see SI), four parameters in particular are poorly constrained in the existing literature: the proportionality constant that modifies the dependence of dissolution rate on runoff and temperature ($k_{d}$), the proportionality constant that modifies the dependence of dissolution rate on runoff only ($k_{w}$), the power constant that modifies the dependence of dissolution rate on the time that a rock particle has spent in the regolith ($\sigma$), and the proportionality constant that modifies the dependence of regolith production on runoff and temperature ($k_{rp}$). As discussed above, the Ca+Mg concentrations of metamorphic and siliciclastic sediment grid cells are also difficult to define. Rather than prescribing values for these six parameters, we instead allow these parameters to vary within reasonable bounds, resulting in a total of 78,000 unique parameter combinations. For each of these combinations, we compute spatially-resolved long-term \COtwo consumption (i.e. Ca+Mg consumption) using present-day runoff, temperature, and slope fields. Following the approach of \citet{Maffre2018a}, we then sum the computed \COtwo consumption over large-scale watersheds that appear in the global compilation of \citet{Gaillardet1999a}, as well as smaller-scale watersheds of the Amazon Basin in the compilation of \citet{Moquet2011a, Moquet2016a, Moquet2018a}. We then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these basins. We eliminate all parameter combinations that result in unreasonable total global \COtwo consumption (which we define to be \textless1\times10$^{12}$~mol/yr or \textgreater1\times10$^{13}$~mol/yr) and/or result in a low $r^{2}$ (which we define to be \textless0.4). This calibration method leaves 3,339 unique parameter combinations that produce global and individual watershed \COtwo consumption fluxes that approximate those estimated in the literature for the present-day.

@@

We use geological and paleoshoreline data to quantify the changes in the area of I\&NG over the past 20~m.y. (Fig. \ref{fig:shoreline_growth}). Following \citet{Molnar2015a}, we analyzed the area changes of islands that are greater than $\sim$200~km$^{2}$, and we also included changes in areas of submerged continental shelves that were previously exposed, like the Sunda Shelf. Carbonate strata and thin-bedded siliciclastic strata were interpreted to represent a subaqueous marine environment, whereas evidence for exposure and terrestrial siliciclastic deposits were used as evidence for a subaerial environment. Taken together, these constraints delineate the position of shorelines. For the island of Sulawesi, we followed the recent stratigraphic compilation and paleoshorelines delineated by \citet{Nugraha2018a} and for New Guinea we followed work of \citet{Norvick2003a, Harrington2017a, Gold2017a}. Although there are difference in data sources and paleoshoreline outlines from \cite{Molnar2015a}, the calculated area at 5~Ma is similar.

Along with the Neogene increase in subaerially-exposed land area and tectonic topography associated with arc-continent collision, there has also been a decrease in subaerially-exposed land area on the Sunda Shelf \citep{Sarr2019a}. Large portions of this low-relief area of continental crust were likely emergent throughout the Neogene, and were repeatedly subaerially exposed during the Pleistocene as recently as the Last Glacial Maximum \citep{Hall2002a}, potentially driving climate amplification on orbital timescales \citep{DiNezio2018a}. Similarities in terrestrial biotas of Borneo and mainland Southeast Asia suggest the existence of land bridges on the Sunda Shelf throughout the Miocene \citep{Moss1998a}. During the Eocene through early Miocene, the rifting of the South China Sea resulted in subsidence of the Sunda Shelf \citep{Morley2013a} and recent subsidence is interpreted to be a result of dynamic topography, triggered by subduction-related mantle flow \citep{Sarr2019a}. In the Miocene, this region exposed a mixture of felsic basement rocks and siliciclastic sediments \citet{Darmadi2007a}. The relatively low-relief and felsic lithologies of the region contrasts with the high-relief and mafic lithologies of most of Indonesia and New Guinea.

@@

I\&NG are potent sinks of carbon because they contain large swaths of ophiolites that are exposed along E-W belts of tectonic topography in the warm wet tropics. In New Guinea, these paleogeographic features originated in the Middle Miocene with the collision between the Melanesian Arc and the distal margin of Australia \citep{vanUfford2005a, Cloos2005a, Baldwin2012a}, which resulted in the exhumation of the 1440~km long Irian-Marum-April Ophiolite Belt. The Middle Miocene (16-14~Ma) basal Markats Formation contains siliciclastic sediment that was transported from the south into the forearc basin associated with the New Guinea ophiolite belt \citep{Cloos2005a}. This precedes, by several million years, the beginning of widespread synorogenic sedimentation to the south and on the Australian continental basement at ca. 12~Ma \citep{vanUfford2005a}. Exhumation of the Central Range accelerated over the past 4~m.y. due to slab-breakoff and buoyant uplift under New Guinea \citep{Cloos2005a}. After arc-continent collision in New Guinea, a left-lateral strike-slip margin developed on the northern margin \citep{Baldwin2012a}. Over the past 4~m.y., this change in plate motion caused the westward extrusion and exhumation of the Cretaceous to Oligocene ophiolites of Obi, Halmahera, and northern New Guinea \citep{Baldwin2012a}. In eastern New Guinea, progressive jamming of the north-dipping subduction zone has caused major uplift over the past 4~m.y. \citep{vanUfford2005a}, which is well dated with a change in provenance from continental to volcanic detritus (Abbott et al., 1994) and thermochronology (Hill & Gleadow, 1989).

@@

However, it is important to acknowledge the limitations of the approach taken in this study. First, the tested scenarios only simulate changes in the location of the shorelines in I\&NG and the lithology exposed in the Deccan and Ethiopian Traps. Changes in paleogeography that are not accounted for in our simulations may have resulted in changes in oceanic and atmospheric circulation \citep{Haug1998a, Shevenell2004a, Holbourn2015a, Molnar2015a} that contributed to changes in \pCOtwo on hundreds of thousands to million-year timescales. However, the global extent and multi-million year timescale of the long-term cooling trend since 14~Ma suggests that the primary driver for this cooling is changes in the geological sources and sinks of carbon, modulated by the silicate weathering feedback \citep{Walker1981a, Raymo1991a, Berner1997a, Kump1997a, Berner2001a}. In this way, changes in the flux of \COtwo outgassing via changes in spreading rates and ridge lengths \citep{Rowan2016a} and changes in weatherability via mountain building and the movement of land masses into/out of areas of high temperature and runoff are likely more important paleogeography-related factors in terms of modulating Neogene cooling than changes in oceanic and atmospheric circulation. Second, the climate model component of GEOCLIM that was used in this study (GFDL CM2.0) has a low spatial resolution (2.0\degrees $\times$ 2.5\degrees) and is based upon physics and geochemical cycles that have been superceded by more recent climate models. Using an updated climate model at higher resolution may result in runoff and temperature fields with different sensitivities to paleogeography and \pCOtwo. Finally, although the silicate weathering component of GEOCLIM now implements lithology-dependent Ca+Mg concentration, it does not yet implement lithology-dependent kinetics of mineral dissolution. Nevertheless, relative to felsic lithologies, mafic lithologies more dominantly comprise of minerals with faster kinetics (e.g. olivine and plagioclase; see SI), and thus incorporating lithology-dependent kinetics of mineral dissolution into our simulations would likely increase the sensitivity of steady-state \pCOtwo to increases in subaerially-exposed land area and tectonic topography of Indonesia and New Guinea.

@@

%%%
% maybe place this at the end of the discussion
The movement of the Deccan Traps into the tropical rain belt together with the low-latitude emplacement of the Ethiopian Traps is an hypothesis that has been put forth to explain the draw-down of \COtwo leading into Oligocene glaciation of Antarctica \citep{Kent2008a, Kent2013a}. To explore the sensitivity of global weatherability and climate state to the presence of the surface exposure of basaltic lavas, we test three additional scenarios using GEOCLIM (Fig. \ref{fig:scenario_pCO2}). In the ``removed Deccan Traps'' and the ``removed Ethiopian Traps'' scenarios, all volcanics associated with the respective igneous provinces are replaced with the Ca+Mg concentration of bulk continental crust. In the ``removed both traps'' scenario, all volcanics associated with both provinces are replaced with the Ca+Mg concentration of bulk continental crust. These scenarios result in steady-state \pCOtwo's \textless500~ppm, distinctly lower than those for the three I\&NG scenarios.
%%%

@@

After the MCO, a cooling trend began ca. 14~Ma that accelerated over the past 4~m.y. Many hypotheses have been proposed to explain this cooling trend including changes in ocean circulation and upwelling \citep{Shevenell2004a, Holbourn2015a}, perhaps associated with increased organic carbon burial \citep{Vincent1985a, Flower1993a} or the closure of the Central American Seaway \citep{Haug1998a}, an increase in the strength of the Walker Circulation \citep{Molnar2015a}, feedbacks with terrestrial vegetation \citep{Cerling1997a}(MAYBE ONE OF THE DINEZIO REFS), uplift in the Himalayas \citep{Raymo1988a}, or a decrease in volcanic outgassing \citep{Berner1983a}(ROWAN+ROWLEY). Although changes in circulation and vegetation may have contributed to changes in \pCOtwo on hundreds of thousands to million-year timescales, the global extent and multi-million year timescale suggests that the long-term cooling trend since 14~Ma is the product of changes in the geological sources and sinks of carbon, modulated by the silicate weathering feedback \citep{Walker1981a, Raymo1991a, Berner1997a, Kump1997a, Berner2001a}. Global spreading rates, ridge length, and paleogeography elsewhere on Earth undoubtedly changed through the Neogene contributing to changes in geological carbon sources and sinks, however, here we show that the global climate state is particularly sensitive to the tropical paleogeograpahy of Indonesia and New Guinea.

@@

Neogene cooling and the emergence of Indonesia and New Guinea

Abstract
The cooling trend from 15 Ma to present coincides with arc-continent collision and exhumation in Indonesia and New Guinea. To assess relationships between the emergence of highly weatherable arc terranes with tectonic topography in the tropics, an increase in global weatherability, geochemical records, and climate change, we use geological and paleoshoreline data quantify the changes in the area of volcanic and ultramafic rocks in Indonesia and New Guinea over the past 20 Myrs. We then incorporate these rock units into a paleogeographic model to determine the position of these units relative to the warm wet tropics. We find that the subaerially exposed area in the tropics of Indonesia and New Guinea has steadily increased through the Neogene with an acceleration over the past 5 Myrs. Arc-continent collision has resulted in the exhumation of both the East Sulawesi and Irian Jaya-New Guinea ophiolites, which are two of the three largest preserved ophiolites on Earth. We suggest that cooling from 15 to 10 Ma was due in large part to the subaerial exposure of these ophiolites and increase in global weatherability, and that the more recent acceleration in cooling was due to increased exhumation rates in Indonesia and New Guinea. Both a decrease of mid ocean ridge spreading rates and changes in the composition and amount of riverine input is necessary to explain both global cooling and geochemical proxy data. The data presented herein helps disentangle these effects.

Significance statement
	The tectonic drivers for the Neogene cooling trend from ~15 Ma to present are unknown. Here we quantify paleoshoreline data in Indonesia and New Guinea over the Neogene to assess the hypothesis that cooling was caused by changes in paleogeography in the warm wet tropics. These data document the emergence of Indonesia over this interval and are consistent with changes in global weatherability associated with arc-continent collision in the  tropics as a major driver of long-term climate change.

\body
After the Miocene Climatic Optimum (MCO), a cooling trend began at ~15 Ma, and accelerated over the past 5 Ma (1-4) (Fig. 1).  Many hypotheses have been proposed to explain this cooling trend including changes in ocean circulation and upwelling (5, 6), perhaps associated with increased organic carbon burial (7, 8) or the closure of the  Central American Seaway (9), an increase in the strength of the Walker Circulation associated with the emergence of Indonesia (10), feedbacks with terrestrial vegetation (11), a decrease in volcanic outgassing (12), an increase in basaltic watersheds in the tropics (10, 13), uplift in the Himalayas (14), and arc-continent collision in the tropics (15). Although changes in circulation and vegetation may have contributed to these signals on hundreds of thousands to million year timescales, the global extent and multi-million year timescale suggests that the long-term cooling trend since 15 Ma is the product of changes in the geological sources and sinks of carbon, modulated by the silicate weathering feedback (16-20).
On geological timescales, atmospheric CO2 levels can vary due to changes in volcanic CO2 flux, but also due to changes in global weatherability — the cumulative factors that affect chemical weathering aside from climate (20, 21). Global weatherability is the product of variables such as lithology, tectonic uplift rates, and paleolatitude (20). An increase in global weatherability can result in cooling as the consumption of carbon through silicate weathering will match that of volcanic input at lower concentrations of atmospheric CO2 (20).
Present-day global CO2-consumption estimates emphasize the importance of highly weatherable areas: ~10-20% of land area is responsible for ~50-75% of CO2 consumption through silicate weathering and associated carbonate precipitation (22, 23). These highly weatherable areas (e.g. Indonesia and New Guinea) are mainly situated in the warm, wet tropics with significant topographic relief, and have bedrock composed of Ca- and Mg-rich mafic and ultramafic lithologies, which have faster dissolution rates than felsic rocks (20, 23, 24). As a result of the intertropical convergence zone (ITCZ), and its migration through the year, annual rainfall on Earth is greatest within ~15º of the equator resulting in the tropical rain belt. Given that hemispherically asymmetric forcings are relatively small, the position of the ITCZ is interpreted to be stable even with large changes in forcing resulting from changing CO2 levels (25). In modern climate data, the region within 10º of either side of the equator corresponds with precipitation greater than 1.5 m/yr and the region within 15º corresponds with precipitation greater than 1.0 m/yr. Because of the strong relationship between runoff and silicate weathering (26), paleogeographic changes in tropics, particularly in tropical regions with tectonic topography and abundant volcanic rocks, like Indonesia and New Guinea, should have an outsized effect on the Earth’s long-term climate state (10, 15).
Recently, it was proposed that arc-continent collision in the tropics drove the cooling trend over the past 15 million years (15, 27). This hypothesis builds off of the suggestion of Reusch and Maasch (28) that arc-continent collisions could accomplish both a decrease in outgassing by terminating arc volcanism (footnote: However, arc-continent collision generally results in subduction flip with renewal of arc magmatism (29)), and an increase in silicate weathering by exhuming soluble Ca- and Mg-rich silicate rocks. Jagoutz et al. (27) applied this process specifically to the closure of the Neo-Tethys ocean in the tropics and suggested that arc-continent collisions at approximately 80 and 50 Ma could account for the long-term, two-phased cooling trend from the Cretaceous to the E-O boundary. Importantly, Jagoutz et al. (27) combined the Himalayan uplift hypothesis (14) with the basalt drift hypothesis (13, 30), emphasizing that topography, precipitation in the tropical rain belt, and lithology are the key variables in global weatherability. Macdonald et al. (15) compiled paleolatitude data on ophiolite bearing sutures through the Phanerozoic and concluded that the extent of arc-continent collisions tropics correlates with extent of continental glaciation, and specifically, that arc-continent collisions in Indonesia and New Guinea were associated with cooling over the past 15 Myr.
Due to rugged tectonic topography (Fig. 2A), tropical temperatures and precipitation rates (Fig. 2B), and a predominance of highly soluble mafic and ultramafic rocks (Fig. 3A), watersheds in Indonesia and New Guinea have the highest rates of silicate weathering in the world accounting for ~9% of the modern global carbon sink (23, 26). Although there are significant monsoonal and orographic effects in Indonesia, and particularly in New Guinea (Fig. 2A), the high mean annual rainfall sill results in rivers with the highest average silicate weathering rates (26). Recognizing the importance of the paleogeography of this region, Molnar and Cronin (10) used paleoshoreline data to estimate that the area of subaerially exposed landmass in Indonesia and New Guinea increased by 60% over the last 5 Myrs. Here we quantify the changes in exposed area and rock types in the tropical region of Indonesia and New Guinea over the past 20 Myrs in a paleogeographic model. These results allow us to assess the changes of mafic and ultramafic rocks in the warm wet tropics of Indonesia and New Guinea throughout the Neogene and test the hypothesis that the subaerial emergence of this region associated with arc-continent collision caused an increase in global weatherability and cooling. These data can further be incorporated into future coupled climate and weathering models to assess the effect of changing paleogeography and global weatherabilty on long-term climate state.

Tectonic Evolution
During Neogene arc-continent collision between the Australian-Indian and Eurasian plates and intervening arc terranes (31-33), Indonesia and New Guinea emerged as several large supra-subduction ophiolites and associated forearcs were exhumed in the warm, wet tropics. Some of these ophiolites formed immediately prior to collision and others have an older history.
The oldest ophiolites preserved in the Indonesia orogenic system are situated in the northwest and associated with the easternmost Paleo-Tethyan sutures (34). On the Malay Peninsula, the Bentong-Raub belt (suture 1; Fig. 3) demarcates the eastern margin of Cimmerian ribbon continent Sibumasu, which collided with the Suhothai arc in the latest Permian to early Triassic (34). Thermochronology suggests that the  Malay Peninsula was first exhumed during the mid-Eocene to early Oligocene (35). This region became a significant sediment source for the Crocker Fan on the NW margin of Borneo during the Oligocene (36), as well as to the east of Malay Peninsula during the Neogene, and has been largely exposed over the last 20 Ma (37, 38).
In Sumatra, the Wolya arc (suture 2; Fig. 3) formed at 95 Ma and is overlain by Campanian 83.5-70.6 Ma radiolarian cherts (39), which were uplifted and eroded from 60-40 Ma (40). These ophiolites were entrained in the Banda Arc, which has experienced magmatism and accretion throughout most of the Cenozoic as the Australian plate subducts beneath it (37, 41). From ~20-10 Ma, the majority of Sumatra was submerged, although Bangka and Belitung were exposed in addition to a portion of the greater Sunda Shelf, which was at least partially exposed throughout the past 20 Ma (37, 41). Most of Sumatra was submerged until after 5 Ma with the arrival of the leading edge of continental Australia (37, 41).
The Meratus belt (sutures 3 and 4; Fig. 3), which extends from Java (42) through eastern Borneo (43) to Palawan (44), marks the Late Cretaceous collision of the Banda blocks and Argo. After this collision, subduction polarity reversed and a Cretaceous to Miocene SE-facing volcanic arc developed (45), which we refer to here as the Great Indonesian arc. In southern Borneo, sedimentological data in the adjacent Barito Basin suggests that ophiolites in the Meratus Mountains were not emergent until the Late Miocene (46). In northern Borneo and Palawan, collision occurred in the Late Oligocene to Early Miocene, resulting in exhumation of the Palawan Ophiolite Complex (44, 46). Over 10 km of Neogene sediment fills the Sarawak Basin and NW Borneo trough, indicative of extensive exhumation and erosion in northern Borneo (36). Thermochronological data suggest extremely rapid uplift (7 km/Myr) and exhumation during the Late Miocene and Early Pliocene, resulting in the recent formation of Mt. Kinabalu, the tallest peak in Borneo (47).
The East Sulawesi ophiolite (suture 5; Fig. 3) was likely generated in a supra-subduction to backarc setting in the Great Indonesian arc (45, 48, 49) [but see (50) for an alternative view], detached from its metamorphic sole in the late Oligocene (32-28 Ma; 51), and subsequently thrust to the east above a west dipping slab. During the Early to Middle Miocene, the Tukang Besi-Buton block began to collide obliquely with the southeastern end of the Sunda Arc, causing additional ophiolite emplacement in west and southeast Sulawesi, including Buton (52, 53). Uplift was diachronous, not effecting units is western Sulawesi until the Middle Miocene (~13 Ma) (54). Exhumation is recorded by the presence of a major Middle Miocene unconformity and sedimentary breccias in marginal basins (52, 54). Collision of the Sula-Banggai block with the east margin of Sulawesi began in the Late Miocene and uplift accelerated in the Early Pliocene (5.2 to 3.8 Ma) and is associated with a major pulse of sedimentation in adjacent basins (55, 56). Off the northeastern margin of the east arm of Sulawesi, Miocene platform carbonates on the Sula-Banggai block are overlain by Late Miocene to Early Pliocene ophiolite detritus in the Celebes mélange (55). Thus, although the East Sulawesi ophiolite was trapped/emplaced onto the composite Sundaland margin (i.e. a fragment of previously accreted Australian crust) between 36 and 23 Ma, mafic and ultramafic rocks appear to have not been substantially subaerially exposed in the south and west arms until after 15 Ma, and on the east arm until after 5 Ma. Thermochronology indicates that exhumation accelerated in the Late Miocene to Pliocene (52-54, 57). North Sulawesi, where active volcanoes extend through the Sangihe Arc into the southern Philippines, was not substantially emergent until the Pliocene (58).
Along the Papuan Peninsula of Eastern New Guinea, Oligocene arc-continent collision between Australian continental fragments and the Inner Melanesian arc exhumed the Papuan Ultramafic Belt (Fig. 3; suture 6) above a north dipping slab (59, 60). In the Central Range of New Guinea, a second set of arc-continent collisions in New Guinea began in the Middle Miocene with the collision between the Outer Melanesian arc and the distal margin of Australia (59, 61, 62). This resulted in the exhumation of the 1440 km long Irian-Marum-April ophiolite belt (Fig. 3; suture 7). The Middle Miocene (16–14 Ma) basal Markats Formation contains siliciclastic sediment that was transported from the south into the forearc basin associated with the PNG ophiolite belt (61, 63). This precedes, by several million years, the beginning of widespread synorogenic sedimentation to the south and on the Australian continental basement at ca. 12 Ma (59). Mountain building began in the Late Miocene, ca. 8-7 Ma (59, 62) with 3-4 km of denudation between 8-5 Ma (64), but major relief was not generated until the Pliocene (65). Exhumation of the Central Range accelerated over the past 5 Myrs due to slab-breakoff and buoyant uplift under New Guinea (61). Although the Papuan Ultramafic Belt was generated and obducted earlier than the ophiolites in the Central Range (60, 66), it was also exhumed very rapidly over the past 10 Ma (62).
After arc-continent collision in New Guinea, a left-lateral strike-slip margin developed on the northern margin (62). Over the past 4 Myr, this change in plate motion caused the westward extrusion and exhumation of the Cretaceous to Oligocene ophiolites of Obi (67), Halmahera (68, 69), and northern New Guinea (70) (Fig. 3; sutures 8-10). In the Bird’s Head region, this change in plate motion led to CCW rotation and the exhumation of several core complexes (71).
The Philippine ophiolites have been grouped in four distinct belts (72). We combine belts 1 and 2 in an eastern suture zone and belts 3 and 4 in a western suture zone (Fig. 3; sutures 11 and 12). Belts 1 and 2 were juxtaposed in the eastern suture zone during sinistral transpression, which resulted in Early Miocene uplift and deposition of coarse clastic sediments (73). The Palawan microcontinental block and Philippine mobile belt collided during late Early Miocene to early Middle Miocene resulting into the emplacement of the western ophiolites, with exhumation extending from the Late Miocene to the present (74).
In addition to collision, arc magmatism may have contributed more significantly to crustal growth in the Philippines (75).
The youngest ophiolites of Indonesia formed in the Banda Sea starting at ca. 15 Ma (33). The ongoing arc-continent collision became emergent in Seram and Timor (Fig. 3; sutures 13 and 14) over the last 5 Myr, as the Banda arc began to abut the Banda embayment emplacing the Timor nappe (49, 76, 77).

Paleoshorelines
Geological maps and paleoshoreline data were used to calculate the changes in different types of subaerially exposed rocks in Indonesia and New Guinea. Following Molnar and Cronin (10), we analyzed the area changes of islands that are greater than ~200 km, and we also included changes in areas of submerged continental shelves that were previously exposed, like the Sunda shelf. Larger islands were further divided into regions based upon their position on microplates as defined by Matthews, et al. (78). Lithological maps were modified from Hartmann and Moosdorf (79) and then supplemented with regional geological maps. From these maps, the percentage of ultramafic and volcanic rocks relative to other rock types was calculated in QGIS for each island and microplate region.
Generally, the paleoshoreline reconstructions within the paleogeographic models coincide with sub-Epoch boundaries, i.e. Early Miocene = 23-16 Ma, Middle Miocene = 15.99-11.61 Ma, Late Miocene = 11.6-5.4 Ma, Pliocene =5.39-3.8 Ma. However, some additional shorelines have been added where additional age information is present to refine the timing of major changes in area. Miocene to present stratigraphic columns were compiled from each region to develop an age model for various paleoenvironmental indicators. In general, carbonate strata and thin-bedded siliciclastic strata were assumed to be deposited in a subaqueous marine environment, whereas evidence for exposure and terrestrial siliciclastic deposits were used as evidence for subaerial environment. Following previous work on paleoshorelines and these environmental indicators, paleoshorelines were outlined in QGIS to calculate areas for the Early Pliocene (5 Ma), Late Miocene (10 Ma), Middle Miocene (15 Ma), and Early Miocene (20 Ma).
We used geological maps and paleoshoreline data to calculate the changes in different types of subaerially exposed rocks in Indonesia and New Guinea. Using Gplates, paleoshoreline shapefiles were incorporated in the paleogeographic model of Matthews, et al. (78). Although there are some differences between this reconstruction and the recent reconstruction of Hall (33), these do not entail significant changes in paleolatitude.
Many new sources for paleoshoreline data have become available since Molnar & Cronin (10). Particularly, for Sulawesi, we followed the recent stratigraphic compilation and paleoshorelines delineated by Nugraha and Hall (80), and in New Guinea, we followed Gold et al. (81),  Cloos et al. (61), Norvick (2003) and derivative work by Harrington et al. (82), and Nichols and Hall (1991). We also compiled our own stratigraphies for the region to provide further paleoenvironmental context. Although the outlines of our paleoshorelines are very different from those of Molnar & Cronin (10) our calculated area for 5 Ma is not substantially different (Fig. X).

>>>>

Of the various parameters used in the equations that govern the silicate weathering component of GEOCLIM (see SI), four parameters in particular are poorly constrained in the existing literature: the proportionality constant that modifies the dependence of dissolution rate on runoff and temperature ($k_{d}$), the proportionality constant that modifies the dependence of dissolution rate on runoff only ($k_{w}$), the power constant that modifies the dependence of dissolution rate on the time that a rock particle has spent in the regolith ($\sigma$), and the proportionality constant that modifies the dependence of regolith production on runoff and temperature ($k_{rp}$). As discussed above, the Ca+Mg concentrations of metamorphic and siliciclastic sediment grid cells are also difficult to define. Rather than prescribing values for these six parameters, we instead allow these parameters to vary within reasonable bounds, resulting in a total of 78,000 unique parameter combinations. For each of these combinations, we compute spatially-resolved long-term \COtwo consumption (i.e. Ca+Mg consumption) using present-day runoff, temperature, and slope fields. Following the approach of \citet{Maffre2018a}, we then sum the computed \COtwo consumption over large-scale watersheds that appear in the global compilation of \citet{Gaillardet1999a}, as well as smaller-scale watersheds of the Amazon Basin in the compilation of \citet{Moquet2011a, Moquet2016a, Moquet2018a}. We then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these basins. We eliminate all parameter combinations that result in unreasonable total global \COtwo consumption (which we define to be \textless1$\times$10$^{12}$~mol/yr or \textgreater1$\times$10$^{13}$~mol/yr) and/or result in a low $r^{2}$ (which we define to be \textless0.4). This calibration method leaves 3,339 unique parameter combinations that produce global and individual watershed \COtwo consumption fluxes that approximate those estimated in the literature for the present-day.

@@

We use geological and paleoshoreline data to quantify the changes in the area of I\&NG over the past 20~m.y. (Fig. \ref{fig:shoreline_growth}). Following \citet{Molnar2015a}, we analyzed the area changes of islands that are greater than $\sim$200~km$^{2}$, and we also included changes in areas of submerged continental shelves that were previously exposed, like the Sunda Shelf. Carbonate strata and thin-bedded siliciclastic strata were interpreted to represent a subaqueous marine environment, whereas evidence for exposure and terrestrial siliciclastic deposits were used as evidence for a subaerial environment. Taken together, these constraints delineate the position of shorelines. For the island of Sulawesi, we followed the recent stratigraphic compilation and paleoshorelines delineated by \citet{Nugraha2018a} and for New Guinea we followed work of \citet{Norwick2003a, Harrington2017a, Gold2017a}. Although there are difference in data sources and paleoshoreline outlines from \cite{Molnar2015a}, the calculated area at 5~Ma is similar.

Along with the Neogene increase in subaerially-exposed land area and tectonic topography associated with arc-continent collision, there has also been a decrease in subaerially-exposed land area on the Sunda Shelf \citep{Sarr2019a}. Large portions of this low-relief area of continental crust were likely emergent throughout the Neogene, and were repeatedly subaerially exposed during the Pleistocene as recently as the Last Glacial Maximum \citep{Hall2002a}, potentially driving climate amplification on orbital timescales \citep{DiNezio2018a}. Similarities in terrestrial biotas of Borneo and mainland Southeast Asia suggest the existence of land bridges on the Sunda Shelf throughout the Miocene \citep{Moss1998a}. During the Eocene through early Miocene, the rifting of the South China Sea resulted in subsidence of the Sunda Shelf \citep{Morley2013a} and recent subsidence is interpreted to be a result of dynamic topography, triggered by subduction-related mantle flow \citep{Sarr2019a}. In the Miocene, this region exposed a mixture of felsic basement rocks and siliciclastic sediments \citet{Darmadi2007a}. The relatively low-relief and felsic lithologies of the region contrasts with the high-relief and mafic lithologies of most of I\&NG.

@@

I\&NG are potent sinks of carbon because they contain large swaths of ophiolites that are exposed along E-W belts of tectonic topography in the warm wet tropics. In New Guinea, these paleogeographic features originated in the Middle Miocene with the collision between the Melanesian Arc and the distal margin of Australia \citep{vanUfford2005a, Cloos2005a, Baldwin2012a}, which resulted in the exhumation of the 1440~km long Irian-Marum-April Ophiolite Belt. The Middle Miocene (16-14~Ma) basal Markats Formation contains siliciclastic sediment that was transported from the south into the forearc basin associated with the New Guinea ophiolite belt \citep{Cloos2005a}. This precedes, by several million years, the beginning of widespread synorogenic sedimentation to the south and on the Australian continental basement at ca. 12~Ma \citep{vanUfford2005a}. Exhumation of the Central Range accelerated over the past 4~m.y. due to slab-breakoff and buoyant uplift under New Guinea \citep{Cloos2005a}. After arc-continent collision in New Guinea, a left-lateral strike-slip margin developed on the northern margin \citep{Baldwin2012a}. Over the past 4~m.y., this change in plate motion caused the westward extrusion and exhumation of the Cretaceous to Oligocene ophiolites of Obi, Halmahera, and northern New Guinea \citep{Baldwin2012a}. In eastern New Guinea, progressive jamming of the north-dipping subduction zone has caused major uplift over the past 4~m.y. \citep{vanUfford2005a}, which is well dated with a change in provenance from continental to volcanic detritus (Abbott et al., 1994) and thermochronology (Hill \& Gleadow, 1989).

@@

The movement of the Deccan Traps into the tropical rain belt together with the low-latitude emplacement of the Ethiopian Traps is an hypothesis that has been put forth to explain the draw-down of \COtwo leading into Oligocene glaciation of Antarctica \citep{Kent2008a, Kent2013a}. To explore the sensitivity of global weatherability and climate state to the presence of the surface exposure of basaltic lavas, we test three additional scenarios using GEOCLIM (Fig. \ref{fig:scenario_pCO2}). In the ``removed Deccan Traps'' and the ``removed Ethiopian Traps'' scenarios, all volcanics associated with the respective igneous provinces are replaced with the Ca+Mg concentration of bulk continental crust. In the ``removed both traps'' scenario, all volcanics associated with both provinces are replaced with the Ca+Mg concentration of bulk continental crust. These scenarios result in steady-state \pCOtwo's \textless500~ppm, distinctly lower than those for the three I\&NG scenarios. This result indicates that, while the presence of high Ca+Mg concentration mafic rocks in low-latitude igneous provinces provides a non-negligible contribution to modulating steady-state \pCOtwo, its contribution is significantly smaller than that of high-relief I\&NG.

However, it is important to acknowledge the limitations of the approach taken in this study. First, the tested scenarios only simulate changes in the location of the shorelines in I\&NG and the lithology exposed in the Deccan and Ethiopian Traps. Changes in paleogeography that are not accounted for in our simulations may have resulted in changes in oceanic and atmospheric circulation \citep{Haug1998a, Shevenell2004a, Holbourn2015a, Molnar2015a} that contributed to changes in \pCOtwo on hundreds of thousands to million-year timescales. However, the global extent and multi-million year timescale of the long-term cooling trend since 14~Ma suggests that the primary driver for this cooling is changes in the geological sources and sinks of carbon, modulated by the silicate weathering feedback \citep{Walker1981a, Raymo1991a, Berner1997a, Kump1997a, Berner2001a}. In this way, changes in the flux of \COtwo outgassing via changes in spreading rates and ridge lengths \citep{Rowan2016a} and changes in weatherability via mountain building and the movement of land masses into/out of areas of high temperature and runoff are likely more important paleogeography-related factors in terms of modulating Neogene cooling than changes in oceanic and atmospheric circulation. Second, the climate model component of GEOCLIM that was used in this study (GFDL CM2.0) has a low spatial resolution (2.0\degrees $\times$ 2.5\degrees) and is based upon physics and geochemical cycles that have been superceded by more recent climate models. Using an updated climate model at higher resolution may result in runoff and temperature fields with different sensitivities to paleogeography and \pCOtwo. Finally, although the silicate weathering component of GEOCLIM now implements lithology-dependent Ca+Mg concentration, it does not yet implement lithology-dependent kinetics of mineral dissolution. Nevertheless, relative to felsic lithologies, mafic lithologies more dominantly comprise of minerals with faster kinetics (e.g. olivine and plagioclase; see SI), and thus incorporating lithology-dependent kinetics of mineral dissolution into our simulations would likely increase the sensitivity of steady-state \pCOtwo to increases in subaerially-exposed land area and tectonic topography of I\&NG.

@@

@@

%After the MCO, a cooling trend began ca. 14~Ma that accelerated over the past 4~m.y. Many hypotheses have been proposed to explain this cooling trend including changes in ocean circulation and upwelling \citep{Shevenell2004a, Holbourn2015a}, perhaps associated with increased organic carbon burial \citep{Vincent1985a, Flower1993a} or the closure of the Central American Seaway \citep{Haug1998a}, an increase in the strength of the Walker Circulation \citep{Molnar2015a}, feedbacks with terrestrial vegetation \citep{Cerling1997a}(MAYBE ONE OF THE DINEZIO REFS), uplift in the Himalayas \citep{Raymo1988a}, or a decrease in volcanic outgassing \citep{Berner1983a}(ROWAN+ROWLEY). Although changes in circulation and vegetation may have contributed to changes in \pCOtwo on hundreds of thousands to million-year timescales, the global extent and multi-million year timescale suggests that the long-term cooling trend since 14~Ma is the product of changes in the geological sources and sinks of carbon, modulated by the silicate weathering feedback \citep{Walker1981a, Raymo1991a, Berner1997a, Kump1997a, Berner2001a}. Global spreading rates, ridge length, and paleogeography elsewhere on Earth undoubtedly changed through the Neogene contributing to changes in geological carbon sources and sinks, however, here we show that the global climate state is particularly sensitive to the tropical paleogeograpahy of I\&NG.

@@
